Treatment heterogeneity in comparative effectiveness of teriparatide vs bisphosphonates: multi-database cohort study

### 2021-02-27 submission/manuscript.rmd
fix typos and rebuild

\newpage
# Methods
## Study design
We ran our analyses on two US observational administrative claims databases and one electronic health record (EHR) database, all mapped to OMOP-CDM version. More specifically, the considered claims databases were IBM MarketScan Medicare Supplemental Database (MDCR) representing retirees in the United States with primary or Medicare supplemental coverage and Optum De-Identified Clinformatics® Data Mart Database – Date of Death (DOD), including members of private health insurance. The EHR database was Optum de-identified Electronic Health Record Dataset (PANTHER) including Humedica’s Electronic Health Record.

Our cohorts included female new users of teriparatide or oral bisphosphonates above the age of 65 (ref 7,8 LEGEND-HTN). Patients were considered to be new-users if they received no anti-osteoporosis drugs (raloxifene, bazedoxifene, denosumab, abaloparatide, romosozumab) 365 days prior to treatment initiation with teriparatide or an oral bisphosphonate. We required patients to have at least one year of observation period before first exposure and an established osteoporosis diagnosis, defined as any recorded fracture of the hip, wrist, spine or shoulder/humerus.

Our primary efficacy outcome was hip fracture. Vertebral fracture and a composite major osteoporotic fracture, defined as hip, vertebral or wrist/forearm/proximal humerus fracture, were our secondary efficacy outcomes. We excluded patients that had experienced the outcome under study any time prior to treatment initiation. Patient time-at-risk started at 1 day after treatment initiation until 730 days after treatment initiation.

##     Statistical analyses
We carried out two sets of analyses. First, we derived overall treatment effect estimates of teriparatide compared to oral bisphosphonates regarding the three outcomes of interest. To account for potential measured confounding we developed separate propensity score models within each database using the same predefined set of measured covariates. Covariates were included in the propensity score model using regularized logistic regression (##REFERENCE Tian-Schuemie simulation). We estimated treatment effects using Cox proportional hazards models with treatment as the sole covariate fitted in the 1-4 propensity score-matched subset of the considered patient population. Results from different databases were summarized using random effects meta-analysis (##REFERECNE).

For the second set of analyses, we used the standardized framework for risk-based assessment of treatment effect heterogeneity, that follows the directives of the PATH statement (##REFERENCE framework). Following data extraction, the framework was applied in two steps. First, we derived individualized risk predictions for the three efficacy outcomes. We built the prediction models using LASSO logistic regression on the propensity score matched (1 to 4) subpopulation of the pooled treatment arms, aiming to remove any effect of measured confounding that may cause differential fit of our models across treatment arms. We considered the same large set of candidate covariates for the development of our prediction models. Time at risk was set at 1 to 730 days after study inclusion.

For each outcome we used the derived prediction models to divide th e population under study in 4 equally-sized subgroups, based on the quarters of the predicted risk distribution. Within each of these risk-based subgroups we developed a new propensity score model. Our analyses were performed on the propensity score matched (1 to 4) subset of the risk subgroup subset. We used a set of XXX negative control outcomes, to calibrate our results for unmeasured confounding (##REFEFRENCE). We derived relative effect estimates using Cox proportional hazards models only with treatment as a predictor. Absolute effect estimates were calculated based on the difference of the Kaplan-Meier estimates.

@@

When assessing heterogeneity of treatment effect, we estimated calibrated hazard ratios of `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(upper)`; 95% CI),
`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(upper)`; 95% CI) and `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)`; 95% CI) within the highest risk quarter of MDCR, OPTUM_DOD and PANTHER respectively. This translated in absolute risk reduction of `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "panther") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(upper)`%; 95% CI). The random effects meta-analytic calibrated hazard ratio within the highest risk quarter was `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(hr)` (`r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(lower)` to `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(upper)`; 95% CI) across the 3 databases.

>>>>

\newpage
# Methods
## Study design
We ran our analyses on three US observational databases mapped to OMOP-CDM version. More specifically:

- **IBM MarketScan Medicare Supplemental Database (MDCR)** is a claims database representing retirees in the United States with primary or Medicare supplemental coverage
- **Optum De-Identified Clinformatics® Data Mart Database – Date of Death (DOD)** is a claims database including members of private health insurance.

- **Optum de-identified Electronic Health Record Dataset (Optum-EHR)** is an EHR database including Humedica’s Electronic Health Record.

Our cohorts included female new users of teriparatide or oral bisphosphonates above the age of 65 (ref 7,8 LEGEND-HTN). Patients were considered to be new-users if they received no anti-osteoporosis drugs (raloxifene, bazedoxifene, denosumab, abaloparatide, romosozumab) 365 days prior to treatment initiation with teriparatide or an oral bisphosphonate. We required patients to have at least one year of observation period before first exposure and an established osteoporosis diagnosis, defined as any recorded fracture of the hip, wrist, spine or shoulder/humerus in their recorded medical history.

Our primary efficacy outcome was hip fracture. Vertebral fracture and a composite major osteoporotic fracture, defined as hip, vertebral or wrist/forearm/proximal humerus fracture, were our secondary efficacy outcomes. We excluded patients that had experienced the outcome under study any time prior to treatment initiation. Patient time-at-risk started 1 day after treatment initiation and finished 730 days after treatment initiation.

##     Statistical analyses
We carried out two sets of analyses. First, we derived overall treatment effect estimates of teriparatide compared to oral bisphosphonates regarding the three outcomes of interest. To account for potential measured confounding we developed separate propensity score models within each database using the same predefined set of measured covariates. Covariates were included in the propensity score model using regularized logistic regression (##REFERENCE Tian-Schuemie simulation). We estimated treatment effects using Cox proportional hazards models with treatment as the sole covariate fitted in the 1-4 propensity score-matched subset of the considered patient population. Results from different databases were summarized using random effects meta-analysis.

For the second set of analyses, we used the standardized framework for risk-based assessment of treatment effect heterogeneity, that follows the directives of the PATH statement (##REFERENCE framework). Following data extraction, the framework was applied in two steps. First, we derived individualized risk predictions for the three efficacy outcomes. We built the prediction models using LASSO logistic regression on the propensity score matched (1 to 4) subpopulation of the pooled treatment arms, aiming to remove any effect of measured confounding that may cause differential fit of our models across treatment arms. We considered the same large set of candidate covariates for the development of our prediction models.

For each outcome we used the derived prediction models to divide the population under study in 4 equally-sized subgroups, based on the quarters of the predicted risk distribution. Within each of these risk-based subgroups we developed a new propensity score model. Our analyses were performed on the propensity score matched (1 to 4) subset of the risk subgroup subset. We used a set of XXX negative control outcomes, to calibrate our results for unmeasured confounding (##REFEFRENCE). We derived relative effect estimates using Cox proportional hazards models only with treatment as a predictor. Absolute effect estimates were calculated based on the difference of the Kaplan-Meier estimates, on day 730 after treatment initiation.

@@

When assessing heterogeneity of treatment effect for hip fractur    e, we estimated calibrated hazard ratios of `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(upper)`; 95% CI),
`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(upper)`; 95% CI) and `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(upper)`; 95% CI) within the highest risk quarter of MDCR, OPTUM_DOD and PANTHER respectively. This translated in absolute risk reduction of `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "panther") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(upper)`%; 95% CI). The random effects meta-analytic calibrated hazard ratio within the highest risk quarter was `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(hr)` (`r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(lower)` to `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(upper)`; 95% CI) across the 3 databases.
